DRAFT \\ From physical principles to classical and quantum \\ Hamiltonian and Lagrangian particle mechanics

### 2016-12-20 2015CompleteDerivation/completeDerivation.tex
Another pass at discussion and conclusions

The first remarkable aspect of this work is that it can exists. Regardless of whether all the minor details are addressed in the best possible way, the idea that so much can be derived from so little is unexpected.

It often happened that the inability to justify one detail forced a partial shift in thinking that enabled to connect more ideas more simply. The ability to constrain the Hamiltonian to vector potential came as a surprise after we realized we needed distributions as a core concept to justify Hamilton's equations. The fact that relationships must be continuous functions came as a surprise after we connected topology with the notion of physical distinguishability. It was this constant simplification that shows us that we were on the right track and that, in the end, decided the boundary of the work as the initial goals were very modest in comparison.

The second striking aspect is a shift in perspective that this work forces upon us. The usual reductionist view is that you start with the rules of the small parts and then build up the composite whole. This works shift that perspective upside down: it's not what the parts can tell you about the whole, it's what the whole can tell you about the parts.

@@

The third notable aspect is that the assumptions have more to do with our ability to create a reliable description rather than essential features of the system itself. A system is not deterministic per se: it is a particular choice for its description (i.e. state) that may be deterministic under some evolution. In the same vein, a material is not homogeneous per se. Or infinitesimally reducible. In this context, the state, its properties and the law of evolution are the description of the system under our idealization, not some intrinsic feature of the system.

The fourth intriguing aspect is that, when framing classical and quantum mechanics in this way, there is a progression. First we assume a homogeneous infinitesimally reducible material. Then we assume an homogeneous irreducible material. Connecting the dots, the next assumption would be a inhomogeneous irreducible material: where the material is not in constant equilibrium and its particles are not composed of the same amount. This different starting point may lead to the resolution of some of the problems we mentioned.

As different particle can transform into one another, they could be seen as different stable states of the same inhomogeneous material. In particular, since $\sigma_x \sigma_v \geq \frac{\hbar}{2m}$, particles in different generations (i.e. electron vs muon) could be similar stable configurations where the fragments of the material are more "packed" in terms of states labeled by the kinematic variables.

@@

In the near future it would seem appropriate to extend this work to the remaining part of well-established physics. The ideas toward that goal are more concrete. The state space for a quantized material, already introduced here, should be captured by the Fock space. The idea of a non-minimal material will force us to consider vector and tensor quantities at each point in the distribution, whose state should be captured by spin systems. Assuming that the vector potential forces themselves are dependent on the state of a material should hopefully give us the Hamiltonian for the fields.

@@

We have seen that much of fundamental physics can be derived from a few simple physical assumption (and lots of hard work). In doing so we obtained a more unified picture of the different branches of fundamental physics centered around the idea of deterministic and reversible evolution of distributions of a homogeneous material. The main physical concepts can be summarized by the following points:

@@

\item Deterministic and reversible evolution is a one-to-one map that preserves the nature of the system (e.g. physical distinguishability, state count and decomposability). Therefore it is an isomorphism on the mathematical structure defined by the state space (e.g. homeomorphism, canonical transformation, unitary transformation), not just a state-to-state bijection.

\item Classical Hamiltonian mechanics describes the evolution of a homogeneous material whose state is equivalent to the states of all its infinitesimal parts, the classical particles. The state space for classical particles is a symplectic manifold $(T^*\mathcal{Q}, \omega)$ as this allows coordinate invariant density distributions; deterministic and reversible evolution for such state space is a symplectomorphism (i.e. a canonical transformation).

@@

\item Quantum (Hamiltonian) mechanics describes the evolution of a homogeneous material for which the state can only be reduced up to finite parts, the quantum particles, whose internal motion is undetermined. The state space for quantum particles is an inner vector product space $(\Psi, \langle \cdot , \cdot \rangle)$; deterministic and reversible evolution for such state space is a unitary transformation.
\end{itemize}

While we are sure that some mathematical details can be improved, we feel that the overall ``conceptual castle" is solid. The starting concepts are simple, the conceptual mapping of those concepts into math is often straight forward and the intuitive ideas that emerge help connecting different branches of math and physics. This is unlikely a matter of simple coincidence.

>>>>

The first remarkable aspect of this work is that something along these lines can be done. Regardless of whether all the minor details are addressed in the best possible way, the idea that so much can be derived from so little was not something we expected. In hindsight, the idea that physics should be founded on physical principles is nothing new: it just hasn't been thoroughly applied to basic mechanics. As these mathematical frameworks have developed across two centuries, it maybe should not be so surprising that they are ripe for some reorganization.

The second striking aspect is a shift in perspective that this work seems to suggest. The usual reductionist view is that you start with the rules of the small parts and then build up the composite whole. This works shift that perspective upside down: it's not what the parts can tell you about the whole, it's what the whole can tell you about the parts.

@@

The third notable aspect is that the assumptions have more to do with our ability to create a reliable description rather than essential features of the system itself. A system is not deterministic per se: it is a particular choice for its description (i.e. state) that may undergo deterministic evolution in some cases. In the same vein, a material is not homogeneous per se. Or infinitesimally reducible. In this context, the state, its properties and the law of evolution are the description of the system under our idealization, not some intrinsic feature of the system.

The fourth intriguing aspect is that, when framing classical and quantum mechanics in this way, there is a progression. First we assume a homogeneous infinitesimally reducible material. Then we assume an homogeneous irreducible material. Connecting the dots, the next assumption would be a inhomogeneous irreducible material: where the material is not in constant equilibrium and its particles are not composed of the same amount. This different starting point may have potential connections to some of the problems we mentioned in the rationales for the assumptions and to other interesting physics questions in general.

As different particles can transform into one another, they could be seen as different stable states of the same inhomogeneous material. In particular, since $\sigma_x \sigma_v \geq \frac{\hbar}{2m}$, particles in different generations (i.e. electron vs muon) could be similar stable configurations where the fragments of the material are more "packed" in terms of states labeled by the kinematic variables.

@@

We limit the discussion to these few remark as the work is already long. We hope to extended it to field theories in the near future. The state space for a quantized material, already introduced here, should be captured by the Fock space. The idea of a non-minimal material will force us to consider vector and tensor quantities at each point in the distribution, whose state should be captured by spin systems. Assuming that the vector potential forces themselves are dependent on the state of a material should hopefully give us the Hamiltonian for the fields. Whether this can be done will be a matter of working through the details, which, we are sure, are likely to bring other unexpected surprises.

@@

We have seen that much of fundamental physics can be derived from few simple physical assumption (and lots of hard work). In doing so we obtained a more unified picture of the different branches of fundamental physics centered around the idea of deterministic and reversible evolution of distributions of homogeneous materials. The main physical concepts can be summarized by the following points:

@@

\item Deterministic and reversible evolution is a one-to-one map that preserves the nature of the system (e.g. physical distinguishability, state count and decomposability). Therefore it is an isomorphism on the mathematical structure defined by the state space (e.g. homeomorphism, canonical transformation, unitary transformation), not just a state-to-state bijection (i.e. isomorphism of sets).

\item Classical Hamiltonian mechanics describes the evolution of a homogeneous material whose state is equivalent to the states of all its infinitesimal parts, the classical particles. The state space for classical particles is a symplectic manifold $(T^*\mathcal{Q}, \omega)$ as this allows to define coordinate invariant density distributions; deterministic and reversible evolution for such state space is a symplectomorphism (i.e. a canonical transformation).

@@

\item Quantum (Hamiltonian) mechanics describes the evolution of a homogeneous material for which the state can only be reduced up to finite parts, the quantum particles, whose internal motion is undetermined. The state space for quantum particles is a complex inner product space $(\Psi, \langle \cdot , \cdot \rangle)$; deterministic and reversible evolution for such state space is a unitary transformation.
\end{itemize}

While we are sure that some mathematical details can be improved, we feel that the overall ``conceptual castle" is solid. The starting concepts are simple, the conceptual mapping of those concepts into math is often straight forward and the intuitive ideas that emerge help to connect different branches of math and physics. This is unlikely a matter of simple coincidence.
